---
title: "Merge all trees"
format: html
editor: visual
---

```{r setup}
library(tidyverse)
library(ape)
library(ggtree)
library(phytools)
```


# MAster list

We have a master list of all marine vertebrates. 

```{r}
a <- readxl::read_xlsx('./Master Species List - Marine Vertebrates.xlsx', guess_max = 10000)
marine_species <- a$species
marine_species <- gsub(pattern = ' ', replacement = '_', marine_species)
```

# Bony fish and sharks/rays

Download the trees:

```{r}
download.file('https://static-content.springer.com/esm/art%3A10.1186%2Fs12862-017-0958-3/MediaObjects/12862_2017_958_MOESM2_ESM.tre', 'betancur_r.tre')
```

```{r}
download.file('https://data.vertlife.org/sharktree/610.tree.10Cal.RAxML.BS.nex', 'sharktree.nex')
```

```{r}
sharks <- ape::read.nexus('sharktree.nex')
write.tree(sharks, 'sharktree.tre')

bonys <- ape::read.tree('betancur_r.tre')
bonys2 <- fishtree::fishtree_phylogeny()
```

Have to rename teh bony fish labels, they have family and G IDs in there.

```{r}
bonys$tip.label <- word(bonys$tip.label, start = 2, end = 3, sep = '_')
write.tree(bonys, 'betancur_r_short_labels.tre')
```

Do we have an intersection of labels? Otherwise we cannot merge

```{r}
intersect(bonys$tip.label, sharks$tip.label)
```

I initially used bind.tree which is blazing fast but just yeets trees onto each other, there's no conflict resoultion. We can't do that.

```{r}
#big_tree <- ape::bind.tree(bonys, sharks)
#big_tree <- ape::bind.tree(big_tree, bonys2)

```

# Adding Aves

```{r}
download.file('https://erda.ku.dk/archives/341f72708302f1d0c461ad616e783b86/B10K/data_upload/03_species_trees/63K.tre', 'aves.tre')
```

```{r}
aves <- read.tree('aves.tre')
```

# Adding Mammalia

```{r}
mammals <- read.tree('./RAxML_bipartitions.result_FIN4_raw_rooted_wBoots_4098mam1out_OK.newick')
mammals$tip.label <- word(mammals$tip.label, start = 1, end = 2 , sep='_')
mammals
```

# Adding reptiles (seasnakes)

sssssss i'm a snake

```{r}
download.file('https://static-content.springer.com/esm/art%3A10.1186%2F1471-2148-13-93/MediaObjects/12862_2013_2346_MOESM1_ESM.txt', 'reptiles.tre')
```

```{r}
reptiles <- read.tree('reptiles.tre')
```

# other vertebrate tree

there's als othis vertebrate tree: https://datadryad.org/dataset/doi:10.5061/dryad.3c6d2

```{r}
download.file('https://dryad-assetstore-merritt-west.s3.us-west-2.amazonaws.com/ark%3A/13030/m5867gh3%7C1%7Cproducer/vertTree.tre?response-content-disposition=attachment%3B%20filename%3DvertTree.tre&X-Amz-Algorithm=AWS4-HMAC-SHA256&X-Amz-Credential=AKIA2KERHV5E3OITXZXC%2F20250331%2Fus-west-2%2Fs3%2Faws4_request&X-Amz-Date=20250331T014242Z&X-Amz-Expires=86400&X-Amz-SignedHeaders=host&X-Amz-Signature=44adc9498fd71c1c2e165d91a3a9f326092f2a1e5df0096a9c15cb781b3d7c07', 'vert.tre')
```

```{r}
merged <- read.tree('vert.tre')
```

The shark tree and the actinopterygii tree have overlapping species so they can be connected, while the Aves and Reptilia tree does not. So I'm using this vertebrate tree as a 'connector'. We'll use this as a 'backbone' so that the Aves tree for example has something to hang onto.


# merging all trees

This takes *DAYS*

```{r}
trees <- c(merged, bonys, bonys2, sharks, mammals, aves, reptiles)
class(trees) <- "multiPhylo"

if(!file.exists('big_tree.Actinopterygii.Chondrichtyes.Mammalia.Aves.Reptilia.tre')) {
    big_tree <- phangorn::superTree(trees, method = 'MRP', multicore = TRUE, mc.cores = 200, trace = 4,
                                rooted = TRUE)
}
```

```{r}
ape::write.tree(phy = big_tree, 'big_tree.Actinopterygii.Chondrichtyes.Mammalia.Aves.Reptilia.tre')
  
```


```{r}
intersect(big_tree$tip.label, bonys$tip.label) %>% length()
intersect(big_tree$tip.label, bonys2$tip.label) %>% length()
intersect(big_tree$tip.label, aves$tip.label) %>% length()
intersect(big_tree$tip.label, reptiles$tip.label) %>% length()
intersect(big_tree$tip.label, sharks$tip.label) %>% length()
intersect(big_tree$tip.label, mammals$tip.label) %>% length()
```

Now we add the marine species not in any of those trees.

First, we merge species into their relative genera:


```{r}
species_in_tree <- big_tree$tip.label %>% unique()

missing_verts <- setdiff(marine_species, species_in_tree)
print(paste0('missing verts: ', length(missing_verts), ' marine verts: ', length(a$species), ' species in tree: ', length(species_in_tree)))
# Conversely, the merged tree has a bunch of non-marine verts - Homo sapiens, Mus musculus, etc.
# those will be useful when we add marine vertebrates!

# now pull out one species per genus from our tree, those we will bind to
species_in_tree_tibble <- tibble(reference = species_in_tree) %>% 
  mutate(genus = word(reference, start =1 ,end = 1, sep = '_')) %>% 
  group_by(genus) %>% 
  slice_head(n=1) %>% ungroup()

dato <- tibble(bind = missing_verts) %>% 
  mutate(bind_g = word(bind, start =1 , end = 1, sep ='_')) %>% 
  left_join(species_in_tree_tibble, by = c('bind_g' = 'genus')) %>% 
    filter(!is.na(reference)) %>% 
  mutate(poly = FALSE)
```

```{r}
genera_in_tree <- species_in_tree %>% word(1, 1, sep ='_') %>% unique()
dato <- dato %>% filter(bind_g %in% genera_in_tree)
```

If we have genera in the dato frame that are not in the tree we're binding to (like *Homo*), then tree.merger() will stop asking us to provide a source tree where that genus should be placed.

We don't have that source tree, so let's remove species we want to add that do not have their genus in the 
species tree. Hopefully as we add more trees we'll be able to place those species.

The following runs for ~5 minutes:


```{r}

# the y-cauda species and y-lineata break everything
big_tree <- ape::drop.tip(big_tree, 'Quietula_y-cauda')
big_tree <- ape::drop.tip(big_tree, 'Starksia_y-lineata')

dato <- dato |> filter(bind != 'Starksia_y-lineata') |> 
  filter(reference != 'Quietula_y-cauda')

if (!file.exists('big_tree.withGeneraAdded.tre')){
  big_tree_merged <- RRphylo::tree.merger(backbone=big_tree, data=as.data.frame(dato %>% select(-bind_g)),plot=FALSE)
}
```

```{r}
ape::write.tree(phy = big_tree_merged, 'big_tree.withGeneraAdded.tre')
```





## Adding species for which the genera are not in the tree

We have about 3600 species where those genera are not in the tree. What we'll do instead
is chuck the species at the root of the family.

So what we have to do:
- get the species that are not in the big tree now, get their genera
- for each genus, get the family from AFD
- calculate the MRCA for all family members that are in the tree
- insert the genus/species at the root of the MRCA


so like this, calculating MRCA of species A and B and inserting there:

```

    --- A
    |
----|
    |
    --- B

```

becomes

```
     --- A
     |
|----|
|    |
|    --- B
|
|------- Added species

```

We will remove branch lengths for this.

```{r}

still_missing <- setdiff(marine_species, big_tree_merged$tip.label)
super_merged3 <- big_tree_merged
super_merged3$edge.length <- NULL # go away branch length
counter <- 1
addeds <- c()
for(spec in still_missing ){
  
  # get all other members in this family
  lineage <- a %>% filter(species == gsub('_', ' ', spec))
  stopifnot(nrow(lineage) > 0)
  this_fam <- lineage$family
  
  all_other_spec_in_family <- a %>% filter(family == this_fam) %>% pull(species) %>% 
    str_replace_all(' ', '_')
  all_other_spec_in_family_in_tree <- intersect(super_merged3$tip.label, all_other_spec_in_family)
  # get the MRCA of all other members
  mrca_spot <- getMRCA(super_merged3, all_other_spec_in_family_in_tree)
  
  # now bind this species to the MRCA
  
  addeds <- c(addeds, spec)
  super_merged3 <- bind.tip(super_merged3, tip.label= spec,
      where=mrca_spot)#, position=max(super_merged$edge.length))
  #super_merged3$edge.length <- c(super_merged$edge.length, max(super_merged$edge.length, na.rm=TRUE))
  counter <- counter + 1
  #if(counter == 5){
  #  break
  #}
}

test_colors <- tibble(label = addeds, source = 'Added into family')

p <- ggtree(super_merged3, branch.length = 'none') %<+% test_colors + geom_tippoint(size=1, aes(color = source)) + scale_color_discrete(na.value='transparent')
ggsave('tree_with_families_added.png', p)
p
```

You can see some at the bottom that are just... bad? Those have been placed at the root of the tree.


```{r}
write.tree(super_merged3, file = 'big_tree.withGeneraAdded.andFamilies.tre')
```

Let's plot the final tree

```{r}
colors <- tibble(label = addeds, source = 'Added later') %>% 
  rbind(tibble(label = setdiff(bonys$tip.label, bonys2$tip.label), source = 'actinops1')) %>% 
  rbind(tibble(label = setdiff(bonys2$tip.label, bonys$tip.label), source = 'actinops2')) %>%
  rbind(tibble(label = intersect(bonys$tip.label, bonys2$tip.label), source = 'actinops1+2')) %>%
  rbind(tibble(label = sharks$tip.label, source = 'sharks')) %>% 
  rbind(tibble(label = reptiles$tip.label, source = 'reptilia')) %>% 
  rbind(tibble(label = aves$tip.label, source = 'aves')) %>% 
  rbind(tibble(label = mammals$tip.label, source = 'mammalia'))
```

```{r}
p <- ggtree(super_merged3, branch.length = 'none') %<+% colors + geom_tippoint(size=1, aes(color = source)) + scale_color_discrete(na.value='transparent')
ggsave('tree_with_families_added.all_sources.png', p)
p
```

